---
title: "Plotting GIS Data (in progress)"
author: "Cole Tanigawa-Lau"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: "flatly"
  html_notebook:
    toc: true
    toc_float: true
    theme: "flatly"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE)
```

# Functions covered:

  - `st_read()`
  - `geom_sf()`
  - `dplyr` function methods for to `sf` objects

# Spatial data and packages in R
GIS data for social science research tend to be stored in one of two formats: vector or raster. Here, we'll be working entirely with the former. Vector data store _features_ such as points, lines, and polygons. These features may represent individual locations (points), roads and rivers (lines), or political units (polygons). Vector data are commonly stored in shapefiles (`.shp`), accompanied by several other files (e.g., `.prj`, `.dbf`) holding accessory data.

<!-- Raster files are common data sources for high-resolution interpolated data, such as rainfall.[^interpolate] Rasters store data in the form of an image with georeferenced pixels, or _cells_. That is, each pixel corresponds to some rectangular physical space, and the value of the data (e.g., amount of rainfall) in that space is the intensity of the corresponding pixel's hue. -->

<!-- [^interpolate]: It would be infeasible to collect rainfall data in all locations of a large gridded area. By observing rainfall in select locations, we can infer the amount of rainfall in unobserved areas, constructing a  -->



<!-- Include stuff about projection/coordinate reference system ? -->


# About `sf`
Standard packages for analyzing spatial data provide methods and object classes specialized for particular types of data. Of course, they also supply functions for reading, writing, generating, and manipulating spatial objects. This tutorial will cover the basic tools in `sf`, a package for representing [simple features](https://en.wikipedia.org/wiki/Simple_Features) in R. Because `sf` is still under development, it does not support several crucial operations, though eventually `sf` will replace the currently standard suite of R spatial packages for vector data: `sp`, `maptools`, `rgdal`, `rgeos`.

Rather than cover the specifics of GIS or how to create simple features from scratch, this tutorial will give a superficial introduction to `sf` objects before demonstrating easy-to-implement manipulation and plotting methods. I learned of the `sf` package from an excellent, but more advanced, [blog post](http://strimas.com/r/tidy-sf/). I recommend going there first, if you're looking to learn more. The package authors include a series of very detailed [vignettes](https://cran.r-project.org/web/packages/sf/). 

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
# If you can't find the function ggplot2::geom_sf(), 
# you may need to download the development version of ggplot2: devtools::install_github("tidyverse/ggplot2")
library(scales)

library(sf)
```

Its package authors designed `sf` to fit neatly into the tidyverse. Like `stringr`, functions begin with a common string: `st_`, an abbreviation of _spatial and temporal_. While the package defines many auxiliary object classes, you will primarily be working with objects that are merely extensions of the basic data frame. 

```{r sf_df}
# Read in precinct-level 2016 presidential election results from Hawaii
pres2016 <- readRDS("data/HI_precinct_gen_pres2016.rds")

# Read in as a data frame
# st_read uses other files in the hi_precinct directory, 
# but you need to point it only to the .shp (shapefile)
hi <- st_read("data/hi_precinct/Precincts_2016_01_13b_PUBLIC.shp")
class(hi)
```

This structure makes indexing, merging, and aggregating spatial objects intuitive; for the most part, you can proceed as you would with any other data frame. What sets `sf` objects apart is their `geometry` column, which holds the GIS data associated with the corresponding row of the data frame. This column follows your `sf` object around as long as it retains its class.[^select_nogeom] To remove the geometry, use `as.data.frame()`.

[^select_nogeom]: Even `select(hi, -geometry)` will return an `sf` object containing a geometry column.

```{r handling}
# Merge with election results data
hi$DP <- as.character(hi$DP)
hi_pres <- left_join(hi, pres2016, by = c(DP = "precinct"))

# Create new variables
hi_pres2 <- mutate(hi_pres,
                   county_code      = as.numeric(gsub("-[0-9]{2}$", "", DP)),
                   total_pres_votes = castle_votes_total + johnson_votes_total + stein_votes_total + 
                                        clinton_votes_total + trump_votes_total,
                   trump_vshare     = trump_votes_total / total_pres_votes,
                   clinton_vshare   = clinton_votes_total / total_pres_votes
                   )
```

```{r, plot}
# Subset to Oahu island
oahu_pres <- filter(hi_pres2,
                    between(county_code, 17, 51))


ggplot(oahu_pres) + geom_sf(aes(fill = clinton_vshare)) + 
  scale_fill_gradient2(midpoint = 0.5, 
                       mid = "maroon4", high = "blue", low = "red", 
                       limits = c(0, 1)) + 
  theme_bw()

# Highlight where third-party votes decided the precinct winner
oahu_pres2 <- mutate(oahu_pres,
                     pwin3rd = abs(clinton_votes_total - trump_votes_total) <= 
                       stein_votes_total + johnson_votes_total + castle_votes_total
                     )
ggplot(oahu_pres2) + geom_sf(aes(fill = pwin3rd)) + theme_bw()

```